---
title: "Rep Data Project 2 (BETTER TITLE NEEDED)"
author: "Gene Kaufman"
date: "January 15, 2016"
output: html_document
---

# Synopsis

Synopsis: Immediately after the title, there should be a synopsis which describes and summarizes your analysis in at most 10 complete sentences.

# Data Processing
First, load some libraries and set some options
```{r setoptions, message=FALSE}
require(knitr)
opts_chunk$set(echo=TRUE, results="asis", warning=FALSE, message=FALSE)

library(dplyr)
library(stringr)
```

Load the file into a data frame
```{r load.data,cache=TRUE}
##### DATA PROCESSING ####
#### Retrieve and load data ####
data_file_url<-"https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
data_file_local<-"storm.data.csv.bz2"
data_file_dir<-"data/"
if (!dir.exists(data_file_dir)) {
  dir.create(data_file_dir)
}
data_file_path<-paste0(data_file_dir,data_file_local)
if (!file.exists(data_file_path)) {
  download.file(data_file_url,data_file_path,mode = "wb")
}
# no need to unzip it - bz2 files can be read directly by read.csv and read.table
if (!exists("data_raw")) {
  data_raw <- read.csv(data_file_path)
}
```

### Prep Data for Analysis
Many entries for **EVTYPE**, **PROPDMGEXP** and **CROPDMGEXP** are duplicated with different cases, set all to upper case
```{r load.data2,cache=TRUE}
data_raw$EVTYPE <- as.factor(toupper(data_raw$EVTYPE))
data_raw$PROPDMGEXP <- as.factor(toupper(data_raw$PROPDMGEXP))
data_raw$CROPDMGEXP <- as.factor(toupper(data_raw$CROPDMGEXP))
```

First, remove all fields, keeping only the ones that we'll need  
Population health:  
- **FATALITIES**, **INJURIES**  
Economic consequences:  
- **PROPDMG**, **CROPDMG** hold dollar amount  
- **PROPDMGEXP**, **CROPDMGEXP** hold character showing magnitude of PROPDMG/CROPDMG
  (i.e. "B","H")
```{r}
wanted_fields <- c("BGN_DATE","STATE","EVTYPE","FATALITIES","INJURIES",
                   "PROPDMG","PROPDMGEXP","CROPDMG","CROPDMGEXP")
data_raw_selected <- data_raw %>%
  select(one_of(wanted_fields))
```

Filter out records that have at least 1 FATALITY, 1 INJURY,
or at least $1 Property or Crop damage
```{r}
data_filtered <- data_raw_selected %>%
  filter(FATALITIES > 0 |
           INJURIES > 0 |
           PROPDMG > 0 |
           CROPDMG > 0)

# data_munged will hold the data as we're reducing the event types
data_munged <- data_filtered
```

Fix a data error in one record - this one was is mentioned in the record comments to be $115M, not $115B  
Credit: Robert Carman in the [https://www.coursera.org/learn/reproducible-research/discussions/DrCe_bl7EeWjxw7W9fJX5Q](Class Discussion forums)
```{r}
data_munged$PROPDMGEXP[data_munged$PROPDMGEXP == 'B' & data_munged$BGN_DATE == '1/1/2006 0:00:00']<- as.factor("M")
```

## Data Munging: Event Type
**Strategy to clean up messy event types:**  
1. Attempt to reduce number of distinct events by removing non-letters, clean up spacing, expand abbreviations and fix obvious misspellings  
2. Explicitly replace events that are mentioned in the NWS documentation as being counted as another event  
3. Loop through official event list, replacing record events if the official event is a substring of the recorded event (e.g. "TORNADO F1" -> "TORNADO")  
```{r}
# some basic replacements to standardize some terms and misspellings
levels(data_munged$EVTYPE) <- gsub("  "," ",levels(data_munged$EVTYPE))
levels(data_munged$EVTYPE) <- gsub("\\\\","/",levels(data_munged$EVTYPE))
levels(data_munged$EVTYPE) <- gsub(" AND ","/",levels(data_munged$EVTYPE))
levels(data_munged$EVTYPE) <- gsub(" FLD$","",levels(data_munged$EVTYPE))
levels(data_munged$EVTYPE) <- gsub("/ ","/",levels(data_munged$EVTYPE))
levels(data_munged$EVTYPE) <- gsub("//","/",levels(data_munged$EVTYPE))
levels(data_munged$EVTYPE) <- gsub("/SML","/SMALL",levels(data_munged$EVTYPE))
levels(data_munged$EVTYPE) <- gsub("/SMALL"," SMALL",levels(data_munged$EVTYPE))
levels(data_munged$EVTYPE) <- gsub("AVALANCE","AVALANCHE",levels(data_munged$EVTYPE))
levels(data_munged$EVTYPE) <- gsub("COASTALSTORM","COASTAL STORM",levels(data_munged$EVTYPE))
levels(data_munged$EVTYPE) <- gsub("HVY","HEAVY",levels(data_munged$EVTYPE))
levels(data_munged$EVTYPE) <- gsub("ICE ON ROAD","ICY ROADS",levels(data_munged$EVTYPE))
levels(data_munged$EVTYPE) <- gsub("ICE ROADS","ICY ROADS",levels(data_munged$EVTYPE))
levels(data_munged$EVTYPE) <- gsub("BLACK ICE","ICY ROADS",levels(data_munged$EVTYPE))
levels(data_munged$EVTYPE) <- gsub("LAKE EFFECT","LAKE-EFFECT",levels(data_munged$EVTYPE))
levels(data_munged$EVTYPE) <- gsub("LIGHTING","LIGHTNING",levels(data_munged$EVTYPE))
levels(data_munged$EVTYPE) <- gsub("LIGNTNING","LIGHTNING",levels(data_munged$EVTYPE))
levels(data_munged$EVTYPE) <- gsub("PRECIP$","PRECIPITATION",levels(data_munged$EVTYPE))
levels(data_munged$EVTYPE) <- gsub("SQUALL$","SQUALLS",levels(data_munged$EVTYPE))
levels(data_munged$EVTYPE) <- gsub("THUDER","THUNDER",levels(data_munged$EVTYPE))
levels(data_munged$EVTYPE) <- gsub("THUNDEER","THUNDER",levels(data_munged$EVTYPE))
levels(data_munged$EVTYPE) <- gsub("THUNDERE","THUNDER",levels(data_munged$EVTYPE))
levels(data_munged$EVTYPE) <- gsub("THUNDERT","THUNDERS",levels(data_munged$EVTYPE))
levels(data_munged$EVTYPE) <- gsub("THUNER","THUNDER",levels(data_munged$EVTYPE))
levels(data_munged$EVTYPE) <- gsub("THUNDERTSORM","THUNDERSTORM",levels(data_munged$EVTYPE))
levels(data_munged$EVTYPE) <- gsub("THUNDERSORM","THUNDERSTORM",levels(data_munged$EVTYPE))
levels(data_munged$EVTYPE) <- gsub("TORNDAO","TORNADO",levels(data_munged$EVTYPE))
levels(data_munged$EVTYPE) <- gsub("TROM","TORM",levels(data_munged$EVTYPE))
levels(data_munged$EVTYPE) <- gsub("TSTM","THUNDERSTORM",levels(data_munged$EVTYPE))
levels(data_munged$EVTYPE) <- gsub("TUNDER","THUNDER",levels(data_munged$EVTYPE))
levels(data_munged$EVTYPE) <- gsub("UNSEASONABLE","UNSEASONABLY",levels(data_munged$EVTYPE))
levels(data_munged$EVTYPE) <- gsub("WINDCHILL","WIND CHILL",levels(data_munged$EVTYPE))
levels(data_munged$EVTYPE) <- gsub("WINDS","WIND",levels(data_munged$EVTYPE))
levels(data_munged$EVTYPE) <- gsub("COLD/WIND","COLD/WIND CHILL",levels(data_munged$EVTYPE))

# Special case - explicitly renamed in NWSI 10-1605
levels(data_munged$EVTYPE) <- gsub("LANDSLIDE","DEBRIS FLOW",levels(data_munged$EVTYPE))

# Special cases - explicitly mentioned in event type description in NWSI 10-1605
data_munged$EVTYPE[grepl("HYPOTHERMIA",data_munged$EVTYPE)]<-as.factor("COLD/WIND CHILL")
data_munged$EVTYPE[grepl("SLIDE",data_munged$EVTYPE)]<-as.factor("DEBRIS FLOW")
data_munged$EVTYPE[grepl("FREEZE",data_munged$EVTYPE)]<-as.factor("FROST/FREEZE")
levels(data_munged$EVTYPE) <- gsub("ASTRONOMICAL HIGH TIDE","STORM SURGE/TIDE",levels(data_munged$EVTYPE))
levels(data_munged$EVTYPE) <- gsub("HIGH TIDES","STORM SURGE/TIDE",levels(data_munged$EVTYPE))
levels(data_munged$EVTYPE) <- gsub("STORM SURGE","STORM SURGE/TIDE",levels(data_munged$EVTYPE))
levels(data_munged$EVTYPE) <- gsub("BEACH EROSION","HIGH SURF",levels(data_munged$EVTYPE))
data_munged$EVTYPE[grepl("SURF",data_munged$EVTYPE)]<-as.factor("HIGH SURF")
levels(data_munged$EVTYPE) <- gsub("DROWNING","RIP CURRENT",levels(data_munged$EVTYPE))
data_munged$EVTYPE[grepl("BURST",data_munged$EVTYPE)]<-as.factor("THUNDERSTORM WIND")
data_munged$EVTYPE[grepl("GUSTNADO",data_munged$EVTYPE)]<-as.factor("THUNDERSTORM WIND")
data_munged$EVTYPE[grepl("LANDSPOUT",data_munged$EVTYPE)]<-as.factor("TORNADO")
data_munged$EVTYPE[grepl("FIRE",data_munged$EVTYPE)]<-as.factor("WILDFIRE")
data_munged$EVTYPE[grepl("WINTRY",data_munged$EVTYPE)]<-as.factor("WINTER WEATHER")
data_munged$EVTYPE[grepl("DAM BREAK",data_munged$EVTYPE)]<-as.factor("FLASH FLOOD")

# temporarily rename MARINE THUNDERSTORM WIND to mtsw so that we don't lose it when
#   we replace THUNDERSTORM in a later step
levels(data_munged$EVTYPE) = c(levels(data_munged$EVTYPE),"mtsw")
data_munged$EVTYPE[grepl("MARINE THUNDERSTORM WIND",data_munged$EVTYPE)]<-as.factor("mtsw")

# this is specifically NOT a thunderstorm, so rename it
data_munged$EVTYPE[grepl("NON[- ]THUNDERSTORM WIND",data_munged$EVTYPE)]<-as.factor("HIGH WIND")

# some special cases where the official event doesn't match up exactly
levels(data_munged$EVTYPE) = c(levels(data_munged$EVTYPE),"HURRICANE (TYPHOON)")
data_munged$EVTYPE[grepl("HURRICANE",data_munged$EVTYPE)]<-as.factor("HURRICANE (TYPHOON)")
data_munged$EVTYPE[grepl("THUNDERSTORM",data_munged$EVTYPE)]<-as.factor("THUNDERSTORM WIND")

# restore MARINE THUNDERSTORM WIND
data_munged$EVTYPE[grepl("mtsw",data_munged$EVTYPE)]<-as.factor("MARINE THUNDERSTORM WIND")

# load master event types
#   Event_Types.txt is a distinct list of Official Events manually pulled out of
#     NWSI 10-1605 8/17/2007, Section 2.1.1
#   https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf
event_type_official <- read.delim("Event_Types.txt",header = FALSE,col.names = "EventType")
event_type_official$EventType <- as.factor(toupper(event_type_official$EventType))

# cycle through the official event list, and rename the recorded events to match
#   the official one IF the official one is a substring of the recorded event
for(i in 1:nrow(event_type_official)) {
  thisone<-levels(event_type_official$EventType)[i]
  levels(data_munged$EVTYPE) = c(levels(data_munged$EVTYPE),thisone)
  data_munged$EVTYPE[grepl(thisone,data_munged$EVTYPE)]<-as.factor(thisone)
}

# now we need take all of the Event Types that didn't map to an official event
#   and combine them into a single event name "_ALL_OTHERS", so that we can
#   gauge the impact of the unmatched data
unmatched<-data.frame(setdiff(data_munged$EVTYPE, event_type_official$EventType))
names(unmatched)<-c("EVTYPE")

levels(data_munged$EVTYPE) = c(levels(data_munged$EVTYPE),"_ALL_OTHERS")
for(i in 1:nrow(unmatched)) {
  thisone<-levels(unmatched$EVTYPE)[i]
  levels(data_munged$EVTYPE)[levels(data_munged$EVTYPE)==thisone] <- "_ALL_OTHERS"
}
```

## Results

